Robotic Bicycle Balance Assistance Reduces Probability of Falling When Subjected to Mechanical Perturbations

### 2024-03-26 main.tex
Added stats results tables and some subcaption figures.

\usepackage{amsmath}
\usepackage[margin=25mm]{geometry}
\usepackage{siunitx}  % use for all units
\usepackage{todonotes}

@@

\abstract{
  Bicycles are unstable at low speeds. We add a controlled steering motor to a
  consumer bicycle that stabilizes the bicycle at low speeds and apply varying
  magnitude external handlebar perturbations to the bicycle while ridden on a
  treadmill. The probability of not recovering from a handlebar perturbation
  decreases when the balance assist is activated.
}

@@

\section{Methods}
%
\subsection{Bicycle}
%
We modified a standard electric bicycle (Gazelle XXX, Dieren, The Netherlands)
with a custom motor in the head tube cable of applying up to
7~\si{\newton\meter} of torque between the head tube and steer tube. A custom
motor controller converts the commanded amperage to applied torque. We measure
the speed of the rear wheel with an encoder (Magura XXX) and measure the roll
rate of the bicycle with a MEMs rate gyroscope. The balance assist control
algorithm is implemented on microprocessor (Teensy).

\subsection{Balance Assist Control}
We use a forward speed \(v\) gain scheduled proportional roll rate positive
feedback controller to stabilize the bicycle. The commanded steer torque
\(T_\delta\) from the steer motor follows
\begin{align}
  T_\delta = k(4.7\si{\meter\per\second} - v)\dot{\phi}
\end{align}
where 4.7~\si{\meter\per\second} is the weave critical speed predicted from the
open loop bicycle rigid rider dynamics. We use \(k=8\) as the low gain value
and \(k=10\) as a high gain value. Scaling the proportional feedback gain
linearly with respect to speed stabilizes the normally unstable weave mode of
the bicycle from speeds of X~\si{\kilo\meter\per\hour} to
X~\si{\kilo\meter\per\hour} as shown in Figure~\ref{fig:eigenvalues}.

@@

\begin{figure}
  \label{fig:eigenvalues}
\end{figure}

\subsection{Perturbations}
%
We apply longitudinal forces to the ends of each handlebar using an adapted
Bump'Em system~\cite{Tan2020} making use of four motors working in tandem. The
four motors are programmed to apply a light force at all times to keep the
ropes taught and to track a commanded force profile using a PID controller
running on a microprocessor (Arduino XXX, Italy). We measure the force applied
by each motor at the handlebar via four inline load cells with a max load of
XX~\si{\newton}. The commanded force profiles are designed to apply an external
pulsive torque to the front assembly (handlebars, forks, wheel) at magnitudes
varying from X to X. The four motors are arranged at the four corners of a 1
meter wide treadmill (make, Company, City) that can reach speeds of
18~\si{\kilo\meter\per\hour}. The general design is described in detail in the
Van De Velde's MSc thesis~\cite{vandeVelde2022}. Our modifications to that
described in the thesis include simplifying the controller with a inexpensive
microcontroller and the use of a simple safety harness.

@@

\begin{align}
  \log\left(\frac{p_{ij}}{1 - p_{ij}}\right) = \Beta + \sum^K_{k=0} \alpha_k x^k_{ij}
\end{align}

@@

>>>>

\usepackage{amsmath}
\usepackage[margin=25mm]{geometry}
\usepackage{siunitx}  % use for all units
\usepackage{subcaption}
\usepackage{todonotes}

@@

\abstract{
  Uncontrolled bicycles are unstable at low speeds. We add a controlled
  steering motor to a consumer bicycle that stabilizes the bicycle at low
  speeds and apply varying magnitude external handlebar perturbations to the
  bicycle while ridden on a treadmill. The probability of not recovering from a
  handlebar perturbation decreases when the balance assist is activated.
}

@@

\section{Methods}
%
\subsection{Bicycle}
%
We modified a standard electric bicycle (Gazelle XXX, Dieren, The Netherlands)
with a custom motor in the head tube capable of applying up to
7~\si{\newton\meter} of torque between the head tube and steer tube. A custom
motor controller converts the commanded torque to applied torque. We measure
the speed of the rear wheel with an encoder (Magura XXX) and measure the body
fixed roll rate of the bicycle with a MEMs rate gyroscope. The balance assist
control algorithm is implemented on microprocessor (Teensy) and data from all
sensors is logged with a CAN bus.

\subsection{Balance Assist Control}
We use a forward speed \(v\) gain scheduled proportional roll rate positive
feedback controller to stabilize the bicycle. The commanded steer torque
\(T_\delta\) from the steer motor follows the control law
\begin{align}
  T_\delta = k(4.7\si{\meter\per\second} - v)\dot{\phi}
\end{align}
where 4.7~\si{\meter\per\second} is the weave critical speed predicted from the
open loop bicycle rigid rider dynamics. We use \(k=8\) as the low gain value
and \(k=10\) as a high gain value. Scaling the proportional feedback gain
linearly with respect to speed stabilizes the normally unstable weave mode of
the bicycle from speeds of X~\si{\kilo\meter\per\hour} to
X~\si{\kilo\meter\per\hour} as shown in Figure~\ref{fig:eigenvalues}.

@@

\begin{figure}
  \centering
  \subcaptionbox{Uncontrolled root locus of eigenvalue components plotted versus speed.}{
    \includegraphics[width=70mm]{example-image-a}
  }
  \subcaptionbox{Controlled root locus of eigenvalue components plotted versus speed.}{
    \includegraphics[width=70mm]{example-image-b}
  }
  \label{fig:eigenvalues}
\end{figure}

\subsection{Perturbations}
%
We apply external longitudinal forces to the ends of each handlebar using an
adapted Bump'Em system~\cite{Tan2020} which is designed with four motors
working in tandem. The four motors are programmed to apply a light force at all
times to keep the ropes taught and to track a commanded force profile using a
PID controller running on a microprocessor (Arduino XXX, Italy). We measure the
force applied by each motor at the handlebar via four inline load cells with a
max load of XX~\si{\newton}. The commanded force profiles are designed to apply
an external pulsive torque to the front assembly (handlebars, forks, wheel) at
magnitudes varying from X to X. The four motors are arranged at the four
corners of a 1~\si{\meter} wide treadmill (make, Company, City) that can reach
speeds of 18~\si{\kilo\meter\per\hour}. The general design is described in
detail in the Van De Velde's MSc thesis~\cite{vandeVelde2022}. Our
modifications to that described in the thesis include simplifying the
controller with a inexpensive microcontroller and the use of a simpler safety
harness.

\begin{figure}
  \centering
  % TODO : scale these pngs to a reasonable size for an A4 sheet
  \subcaptionbox{}{\includegraphics[width=70mm]{figures/participant-in-set-up.png}}
  \subcaptionbox{Diagram of perturbation system.}{\includegraphics[width=70mm]{figures/setup-diagram.png}
}
\end{figure}

@@

\begin{align}
  \log \left(\frac{p_{ij}}{1-p_{ij}} \right) = \beta + \sum_{k=0}^{K} \alpha_k x_{ij}^{k}
\end{align}

@@

The coefficient estimates for a single-level logistic regression at a speed of 6
\unit{\kilo\meter\per\hour} and gain $k=-10$ are shown in table \ref{table:freq-coefs-6}. The
associated multiplicative change in odds are shown in table \ref{table:freq-change-in-odds-6}.

\begin{table}
  \centering
  \caption{Coefficient estimates at 6 \unit{\kilo\meter\per\hour} and gain
  $k=-10$ based on frequentist single-level logistic regression.}
  \label{table:freq-coefs-6}
  \begin{tabular}{lllll}
    \textbf{Variable}                             & \textbf{Estimate} & \textbf{Standard error} & \textbf{p-value} & \\ \cline{1-4}
    Intercept                                     & -0.29             & 0.17                    & 0.09             & \\
    Angular impulse                               & 1.69              & 0.27                    & 0.00             & \\
    Perturbation order                            & -0.77             & 0.22                    & 0.00             & \\
    Balance-assist on                             & -0.64             & 0.27                    & 0.02             & \\
    Roll angle                                    & -0.25             & 0.21                    & 0.24             & \\
    Steer angle                                   & -0.14             & 0.21                    & 0.51             & \\
    Balance-assist on $\times$ roll angle         & 0.52              & 0.34                    & 0.12             & \\
    Balance-assist on $\times$ steer angle        & -0.41             & 0.34                    & 0.22             & \\
    Balance-assist on $\times$ angular impulse    & 0.41              & 0.41                    & 0.32             & \\
    Balance-assist on $\times$ perturbation order & -0.53             & 0.34                    & 0.12             & \\
  \end{tabular}
\end{table}

\begin{table}
  \centering
  \caption{Multiplicative change in odds of predictor variables at 6
  \unit{\kilo\meter\per\hour} and gain $k=-10$ based on frequentist
         single-level logistic regression.}
  \label{table:freq-change-in-odds-6}
  \begin{tabular}{lllll}
      \textbf{Variable}                             & \textbf{Multiplicative change in odds} & \textbf{2.5\%} & \textbf{97.5\%} & \textbf{p-value} \\ \hline
      Intercept                                     & 0.75                                   & 0.53           & 1.05            & 0.09             \\
      Angular impulse                               & 5.40                                   & 3.18           & 9.16            & 0.00             \\
      Perturbation order                            & 0.46                                   & 0.30           & 0.72            & 0.00             \\
      Balance-assist on                             & 0.53                                   & 0.31           & 0.89            & 0.02             \\
      Roll angle                                    & 0.78                                   & 0.51           & 1.18            & 0.24             \\
      Steer angle                                   & 0.87                                   & 0.58           & 1.32            & 0.51             \\
      Balance-assist on $\times$ roll angle         & 1.68                                   & 0.86           & 3.29            & 0.12             \\
      Balance-assist on $\times$ steer angle        & 0.66                                   & 0.34           & 1.28            & 0.22             \\
      Balance-assist on $\times$ angular impulse    & 1.51                                   & 0.67           & 3.38            & 0.32             \\
      Balance-assist on $\times$ perturbation order & 0.59                                   & 0.30           & 1.15            & 0.12             \\
  \end{tabular}
\end{table}

The coefficients estimates of a single-level logistic regression are shown in table
\ref{table:freq-coefs}. The associated multiplicative changes in odds are shown in table
\ref{table:freq-change-in-odds}.

\begin{table}
  \centering
  \caption{Coefficient estimates at 10 \unit{\kilo\meter\per\hour} based on frequentist single-level logistic
      regression.} \label{table:freq-coefs}
  \begin{tabular}{lllll}
      \textbf{Variable}                             & \textbf{Estimate} & \textbf{Standard error} & \textbf{p-value} & \\ \cline{1-4}
      Intercept                                     & -0.24             & 0.16                    & 0.13             & \\
      Angular impulse                               & 2.39              & 0.29                    & 0.00             & \\
      Perturbation order                            & -1.16             & 0.21                    & 0.00             & \\
      Balance-assist on                             & -0.44             & 0.24                    & 0.07             & \\
      Roll angle                                    & 0.27              & 0.22                    & 0.22             & \\
      Steer angle                                   & -0.37             & 0.24                    & 0.12             & \\
      Balance-assist on $\times$ roll angle         & -0.61             & 0.34                    & 0.08             & \\
      Balance-assist on $\times$ steer angle        & 0.56              & 0.35                    & 0.11             & \\
      Balance-assist on $\times$ angular impulse    & 0.46              & 0.44                    & 0.29             & \\
      Balance-assist on $\times$ perturbation order & -0.37             & 0.32                    & 0.25             & \\
  \end{tabular}
\end{table}

\begin{table}
  \centering
  \caption{Multiplicative change in odds of predictor variables at 10 \unit{\kilo\meter\per\hour} based on
      frequentist single-level logistic regression.} \label{table:freq-change-in-odds}
  \begin{tabular}{lllll}
    \textbf{Variable}                             & \textbf{Multiplicative change in odds} & \textbf{2.5\%} & \textbf{97.5\%} & \textbf{p-value} \\ \hline
    Intercept                                     & 0.78                                   & 0.57           & 1.07            & 0.12             \\
    Angular impulse                               & 10.92                                  & 6.23           & 19.13           & 0.00             \\
    Perturbation order                            & 0.31                                   & 0.21           & 0.48            & 0.00             \\
    Balance-assist on                             & 0.64                                   & 0.40           & 1.03            & 0.07             \\
    Roll angle                                    & 1.31                                   & 0.85           & 2.04            & 0.22             \\
    Steer angle                                   & 0.69                                   & 0.43           & 1.10            & 0.12             \\
    Balance-assist on $\times$ roll angle         & 0.54                                   & 0.28           & 1.07            & 0.08             \\
    Balance-assist on $\times$ steer angle        & 1.76                                   & 0.88           & 3.50            & 0.11             \\
    Balance-assist on $\times$ angular impulse    & 1.59                                   & 0.68           & 3.74            & 0.29             \\
    Balance-assist on $\times$ perturbation order & 0.69                                   & 0.37           & 1.30            & 0.25             \\
  \end{tabular}
\end{table}
